\section{External memory mergesort}
In the external-memory model (hereafter EM model), show how to implement the $k$-way merge (where $(k + 1)B \leq M$), namely, how to simultaneously merge $k$ sorted sequences of total length $N$, with an I/O cost of $O(N/B)$ where $B$ is the block transfer size. Also, try to minimize and analyze the CPU time cost.

\subsection{First solution}

We keep in main memory $k$ input buffers $B_1, B_2, \dots, B_k$ --- one for each sorted sequence $s_1, s_2, ..., s_k$ that we need to merge --- and 1 output buffer $B_{out}$. At each step we find the smallest unchosen element among the input buffers and copy this element to $B_{out}$. When $B_{out}$ is full, we write it to the end of a file $F$, then we clear $B_{out}$'s content.

For each input buffer we store a pointer $p_i$ to its first unchosen element. When a pointer $p_i$ reaches the end of the buffer $B_i$, we need to load in $B_i$ the next portion of the corresponding sorted sequence $s_i$ and reset $p_i$. We also store a pointer to the next free slot in $B_{out}$.

\paragraph{I/O complexity.} The $k$-way merge ends when we reach the end of every sorted sequence: we performed $\Theta(N/B)$ read operations for the sorted sequences, and $\Theta(N/B)$ writes to $F$, thus the total I/O complexity is $\Theta(N/B)$.

\paragraph{Minimizing CPU time.} The CPU complexity of searching the smallest element among the unchosen elements of the input buffers is $O(k)$. We can improve the algorithm by replacing the linear search with a minimum priority queue of the unchosen elements: in this way we can extract the smallest element in $O(1)$, advance the pointer $p_i$ of the corresponding input buffer $B_i$ and finally insert $B_i[p_i]$ in the priority queue in $O(\log k)$ time.

\subsection{Second solution}
Let
    \begin{itemize}
    \item $k = \frac{1}{4}\cdot \frac{M}{B}$ be the number of blocks the cache can hold.
    \item $B_s$ the blocks reserved in cache to store temporary merges.
    \item $N$ be the size of the problem.
    \item $B$ the block size, with $B_i$ a pointer to the $i^{th}$ block.
    \item $B_{i}^{c}$ the $i^{th}$ block stored in memory.
    \item $\frac{N}{B}$ the number of blocks storing $A$ in main memory.
    \end{itemize}
We assume the sorted blocks we receive in input are ordered and use $\frac{1}{3} \cdot \frac{M}{B}$ to store loaded blocks in a Fibonacci min-heap fashion.
\emph{k-way} mergesort operates on $k$ sorted blocks we load in cache.
We implement the merge operation as follows, outputting to external memory on file $O$:
    \begin{algorithmic}[1]
    \Function{merge}{$a$, $b$}:
    \For{$i \gets 0; i < k; i++$}
        \State load($B_i$, $B_i^m$)\;           \Comment{Load blocks in cache: $O(\frac{N}{B})$}
    \EndFor
    \State $busy\_blocks = k - 1$\;             \Comment{\parbox[t]{.5\linewidth}{Count busy blocks, that is blocks whose pointers have not reached
                                                        the end of the block}}
    \State $S \gets \&S$\;                      \Comment{\parbox[t]{.5\linewidth}{At most $O(k \cdot B)$, when every sorted block is greater than the following}}
    \While{$busy\_blocks > 0$}
        \State $m \gets \min\{\min{B_i^c}\}$\;  \Comment{Get the current minimum of every block in cache: $O(k)$}
        \State $B_j^c \gets B_j^c + 1$\;        \Comment{Shift to next element in block $B_j$ with current minimum element}
        \If{$B_j == EOB$}
            \State $busy\_blocks \gets busy\_blocks - 1$\;
        \EndIf
        \If{$busy\_blocks == 1$}
            \State $S \gets S :: flush(B_j^c)$\;    \Comment{Flush remaining block}
        \EndIf
        \State $*S \gets m$\;
        \State $S \gets S + 1$\;                \Comment{Increment storing blocks pointer}
    \EndWhile
    \State write(S, O)\;                        \Comment{Output}
    \EndFunction
\end{algorithmic}

As we assumed, our blocks are ordered Fibonacci min-heaps, thus allowing us to merge them in $k \cdot \log(n)$ time.
We can then store $S$ by finding the minimum element $n$ times in costant time: $n \cdot O(1) = O(n)$, giving us a total cost of
    \begin{equation*}
    k \cdot B + \log(k) \cdot O(k) + O(n)
    \end{equation*}
respectively for construction, find the minimum over $k$ Fibonacchi heaps in the \emph{merge} procedure and store it in memory.
